Predicting _C. difficile_ Infection Severity from the Taxonomic Composition of the Gut Microbiome

### 2023-06-02 paper/paper.qmd
Use mean not median for balanced precision...
WIP: feature importance results

We performed permutation feature importance to determine which OTUs contributed
the most to model performance (@fig-features).
An OTU was considered important if performance decreased when it was permuted in
at least 75% of the train/test splits.

@@

It's not enough for models to perform well to justify deploying them in a
clinical setting; benefit over current practices must be shown.
Estimating the NNB contextualizes model performance within clinical reality.
Amplicon sequencing is not typically performed for CDI patients, but if there is
clinical value to be gained by implementing OTU-based models, routinely
sequencing and profiling the microbial communities of CDI patients could be
justified.

@@

::: {#fig-features fig-cap="**Feature importance.**
**A)** Feature importance via permutation test.
For each OTU, the order of samples was randomized in the test set 100 times and
the performance was re-calculated to estimate the permutation performance.
An OTU was considered important if the performance decreased when the OTU was
permuted in at least 75% of the models.
OTUs with a greater difference in AUROC (actual performance minus permutation
performance) are more important.
Left: models were trained on the full datasets, with different numbers of
samples available for each severity definition.
Right: models were trained on the intersection of samples with all labels
available for each definition. Note that Attributable and Pragmatic severity are
exactly the same for the intersection dataset.
*Pseudomonas* (OTU 120) is not shown for IDSA severity in the full datasets nor
in the intersection dataset because it was removed during pre-processing due to
having near-zero variance.
**B)** Log~10~-transformed median relative abundances of the most important OTUs
on the full datasets, grouped by severity (shape). 
The vertical dashed line is the limit of detection.
"}
![](figures/feature-importance.png)
:::

>>>>

We performed permutation feature importance to determine which OTUs contributed
the most to model performance (@fig-features).
An OTU was considered important if performance decreased when it was permuted in
at least 75% of the train/test splits, with greater differences in AUROC meaning
greater importance.
We plotted log~10~-transformed median relative abundances for the most important
OTUs.
_Enterococcus_ was the most important OTU, being significantly important for 
all models except for attributable severity on the full dataset.
The relative abundance of _Enterococcus_ was also elevated in severe cases over
not severe cases for all severity definitions.
_Staphylococcus_ was important for the pragmatic and all-cause definitions on
the full datasets.
_Lactobacillus_ was important only for the all-cause definition on the intersection
dataset.
All remaining OTUs had differences in AUROC < 0.02 and were only significantly
important in one or two of the models.
For many of the top OTUs, there is wide variance in importance, likely due to
the imbalanced nature of the severity outcomes.
There is not always a clear pattern
of increased or decreased relative abundance of important OTUs in severe cases.

TODO reference other literature?

@@

It's not enough for models to perform well to justify deploying them in a
clinical setting; benefit over current practices must be shown.
do no harm [@wiens_do_2019].
Estimating the NNB contextualizes model performance within clinical reality.
Amplicon sequencing is not typically performed for CDI patients, but if there is
clinical value to be gained by implementing OTU-based models, routinely
sequencing and profiling the microbial communities of CDI patients could be
justified.

@@

::: {#fig-features fig-cap="**Feature importance.**
**A)** Feature importance via permutation test.
For each OTU, the order of samples was randomized in the test set 100 times and
the performance was re-calculated to estimate the permutation performance.
An OTU was considered important if the performance decreased when the OTU was
permuted in at least 75% of the models.
Median difference in AUROC is reported for each OTU, with starred OTUs being
significant for a 75% confidence interval.
OTUs with a greater difference in AUROC (actual performance minus permutation
performance) are more important.
Left: models were trained on the full datasets, with different numbers of
samples available for each severity definition.
Right: models were trained on the intersection of samples with all labels
available for each definition. Note that Attributable and Pragmatic severity are
exactly the same for the intersection dataset.
*Pseudomonas* (OTU 120) is not shown for IDSA severity in the full datasets nor
in the intersection dataset because it was removed during pre-processing due to
having near-zero variance.
**B)** Log~10~-transformed median relative abundances of the most important OTUs
on the full datasets, grouped by severity (shape). 
The vertical dashed line is the limit of detection.
"}
![](figures/feature-importance.png)
:::
